---
title: "Statistical tests"
output: rmarkdown::html_vignette
author: Tomasz Włodarczyk
vignette: >
  %\VignetteIndexEntry{Statistical tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chi-square test of the aggression frequency between colonies nesting within and outside wood ant territory

```{r message = FALSE}
library(fusca.defence)
# Load data with the results of the aggression test - basic variant
data("aggression_probing_tests")
aggression_probing_tests <- aggression_probing_tests[!(aggression_probing_tests$test_ID %in% id_remove),]
# For experiments with the use of 20 F. sanguinea ants introduced into arena around F. fusca nest
# T - within territory, B - beyond territory
# Colonies are classified according to the maximal number of aggressive ants exceeding
# threshold number
chi_sq_test(treatment_ = "sanguinea_20")
```
```{r}
# The same as above but for treatment with the use of a signle F. sanguinea ant
chi_sq_test(treatment_ = "sanguinea_1")
```

# Wilcoxon test to examine difference in duration of aggression displayed by agitated F. fusca ants toward alien ant

```{r message = FALSE}
# Load dataset
data("agitation_tests")
# summarize duration of aggressive behaviour over the experiment
agitation_tests_summary <- group_by(agitation_tests, colony, treatment) %>%
  summarise(aggression_duration=sum(aggression_duration)) %>% 
  left_join(distinct(select(agitation_tests, colony, territory)))
# difference in aggression toward F. polyctena between colonies located within and 
# outside rufa territory
x<-filter(agitation_tests_summary, treatment=="F. polyctena", territory=="B")$aggression_duration
y<-filter(agitation_tests_summary, treatment=="F. polyctena", territory=="T")$aggression_duration
cat(sprintf("Median time of aggression of colonies from within and from outside
wood ant territory: %f sec., %f sec., respectively", median(y), median(x)))
wilcox.test(x,y)
# difference in aggression toward F. sanguinea between colonies located within and 
# outside rufa territory
x<-filter(agitation_tests_summary, treatment=="F. sanguinea", territory=="B")$aggression_duration
y<-filter(agitation_tests_summary, treatment=="F. sanguinea", territory=="T")$aggression_duration
cat(sprintf("Median time of aggression of colonies from within and from outside
wood ant territory: %f sec., %f sec., respectively", median(y), median(x)))
wilcox.test(x,y)
# difference in aggression toward F. sanguinea and F. polyctena outside territory
sang_res <- filter(agitation_tests_summary, treatment=="F. sanguinea", territory=="B") %>%
  select(colony, aggression_duration)
polyctena_res <- filter(agitation_tests_summary, treatment=="F. polyctena", territory=="B") %>%
  select(colony, aggression_duration)
res <- left_join(sang_res, polyctena_res, by=c("colony"))
cat(sprintf("Median time of aggression toward F. polyctena and 
F. sanguinea: %f sec., %f sec., respectively", 
median(res$aggression_duration.y), 
median(res$aggression_duration.x)))
wilcox.test(res$aggression_duration.x, res$aggression_duration.y, paired=TRUE)
```

# Dependence of F. fusca number on the ambient temperature, colony location and experiment duration
```{r}
library(lme4)
library(lmerTest)
library(DHARMa)
model_data <-transform_to_single_val(raw_data = aggression_probing_tests, column = "actual_fus_numb")
# treatment sanguinea_20 should be removed due to significantly lower fusca activity
model_data<-filter(model_data, territory %in% c("B","T"), 
                   date > as.Date("2016-12-30"),
                   date < as.Date("2020-12-30"),
                   !is.na(temperature), 
                   treatment!="sanguinea_20")
aggression_records <- filter(model_data, aggression_num > 0)
new_dataset<-data.frame()
for(i in 1:nrow(model_data)){
  Test = model_data$test_ID[i]
  interval = model_data$time_elapsed[i]
  # include experiments with no aggression
  if (!(Test %in% aggression_records$test_ID)){
    new_dataset <- rbind(new_dataset, model_data[i,])
    next
  }
  # include experiment part before aggression occurred
  reference_set <- filter(aggression_records, test_ID == Test)
  if (all(interval < reference_set$time))
    new_dataset<-rbind(new_dataset, model_data[i,])
}
model_data <- new_dataset
# temperature transformation: center at zero
model_data$temperature <- model_data$temperature - min(model_data$temperature) + 1
# temperature transformation: normalize standard deviation
sd_temperature <- sd(model_data$temperature)
model_data$temperature <- model_data$temperature/sd_temperature
# fit generalized linear model; use log10 transform of the temperature
mixed.model.activity<-glmer(actual_fus_numb ~ temperature + (1|colony) + territory +
                          I(log10(time_elapsed)), data=model_data, family="poisson")
summary(mixed.model.activity)
```

Show diagnostic plots using DHARMa package.
```{r fig.dim = c(9, 6)}
set.seed(1010)
sim.resid <- simulateResiduals(mixed.model.activity, n = 1000)
plot(sim.resid)
```
Retrieve the effect of temperature increase on F. fusca number and back transform into the original scale.
```{r}
temp.effect <- fixef(mixed.model.activity)['temperature']
temp.effect <- temp.effect/sd_temperature
exp(temp.effect)
```
# Logistic model for the probability of F. sanguinea being attacked by F. fusca colony
```{r}
# compare treatments with the use of 1 nad 20 F. sanguinea workers
logistic.model.res <- filter(aggression_probing_tests, date>as.Date("2019-01-01"), territory %in% (c("B", "T"))) %>%
  transform_to_single_val() %>% 
  split(.$test_ID) %>%
  purrr::map(process_test_data) %>%
  do.call(rbind, .) %>%
  filter(is.finite(ant_number)) %>%
  mutate(aggression_presence = ant_number > 0) %>% 
  glmer(aggression_presence ~ treatment + territory + (1|colony), data = ., family = "binomial") 
summary(logistic.model.res) 
```
Create diagnostic plots with the use of DHARMa package 
```{r fig.dim = c(9, 6)}
plot(simulateResiduals(logistic.model.res, n = 1000))
```

# Factors influencing the maximal number of F. fusca ants on the arena
```{r eval = FALSE}
set.seed(1001)
max_activ_data <- data.frame()
max_ac_model_list <-list()
for (i in seq_len(1E3)){
  sample_data <- aggression_probing_tests %>% 
    filter(date > as.Date("2019-01-01")) %>%
    transform_to_single_val(column = "actual_fus_numb", mode = 4) %>% # select random fusca number from observed range
    group_by(territory, treatment, colony, date) %>% 
    summarise(max_activity = max(na.omit(actual_fus_numb)))
  ma_model <- try(glmer(max_activity ~ treatment * territory + (1|colony),
                    data = sample_data, family = poisson(link="log")))
  if (class(ma_model)=="try-error"){
    print("Model error")
    next
  }
  max_ac_model_list[[length(max_ac_model_list)+1]] <- list()
  sim_residuals <-simulateResiduals(ma_model)
  diagnostic_tests <- list(testUniformity(sim_residuals, plot = FALSE)$p.value,
                           testOutliers(sim_residuals, plot = FALSE)$p.value,
                           testDispersion(sim_residuals, plot = FALSE)$p.value,
                           testQuantiles(sim_residuals, plot = FALSE)$p.value)
  names(diagnostic_tests) <- c("testUniformity", "testOutliers", "testDispersion", "testQuantiles")
  max_ac_model_list[[length(max_ac_model_list)]]$glmm_model <- ma_model
  max_ac_model_list[[length(max_ac_model_list)]]$diagnostic_tests <- diagnostic_tests
}
# which model pass all diagnostic tests
max_activ_data <- data.frame()
for(i in seq_along(max_ac_model_list)){
  max_activ_data <- rbind(max_activ_data, data.frame(c(fixef(max_ac_model_list[[i]]$glmm_model), 
                                                       max_ac_model_list[[i]]$diagnostic_tests)))
}
diagnostic_pass <- apply(max_activ_data, 1, function(x) all(x[5:8]>0.05))
cat(sprintf("Number of models which passed all diagnostic tests: %d\n", 
            sum(diagnostic_pass, na.rm = TRUE)))
```
```{r echo = FALSE}
max_activ_data <- data.frame()
max_ac_model_list <- fusca.defence:::max_ac_model_list
for(i in seq_along(max_ac_model_list)){
  max_activ_data <- rbind(max_activ_data, data.frame(c(fixef(max_ac_model_list[[i]]$glmm_model), 
                                                       max_ac_model_list[[i]]$diagnostic_tests)))
}
# which model pass all diagnostic tests
diagnostic_pass <- apply(max_activ_data, 1, function(x) all(x[5:8]>0.05))
cat(sprintf("Number of models which passed all diagnostic tests: %d\n", 
            sum(diagnostic_pass, na.rm = TRUE)))
```
p-values of model coefficients for models which passed diagnostic tests
```{r eval = FALSE}
model_p_vals <- data.frame()
for (i in seq_len(length(max_ac_model_list))){
  model_p_vals <- rbind(model_p_vals, data.frame(as.list(summary(max_ac_model_list[[i]]$glmm_model)$coeff[,4])))
}
p_vals_splitted <- split(model_p_vals, diagnostic_pass) %>% lapply(function(x) apply(x,2,summary))
p_vals_splitted$'TRUE'
```
```{r echo = FALSE}
model_p_vals <- data.frame()
for(i in seq_along(max_ac_model_list)){
model_p_vals <- rbind(model_p_vals, max_ac_model_list[[i]]$p_val) 
}
p_vals_splitted <- split(model_p_vals, diagnostic_pass) %>% lapply(function(x) apply(x,2,summary))
p_vals_splitted$'TRUE'
```
p-values of model coefficients for the rest of the models
```{r}
p_vals_splitted$'FALSE'
```
# Difference in the head width of foragers from colonies inhabiting wood ant territory and those living elsewhere
```{r}
data(head_width)
hw_summary <- headwidth_summary(head_width)
t.test.data <- filter(aggression_probing_tests, territory %in% c("T","B")) %>%
 select(colony,territory) %>% 
  left_join(hw_summary) %>% 
  select(colony,territory,mean) %>% 
  distinct()
x <- filter(t.test.data,territory=="T") %>% pull(mean)
y <- filter(t.test.data,territory=="B") %>% pull(mean)
cat(sprintf("Mead head width of foragers from wood ant territory: %.2f μm", mean(x, na.rm = TRUE)))
cat(sprintf("Mead head width of foragers from outside wood ant territory: %.2f μm", mean(y, na.rm = TRUE)))
t.test(x,y)
```
